Brownian local times

### 2015-10-05 notes/brownian-local-times.Rmd
added html

It states in [Takacs](https://projecteuclid.org/euclid.aoap/1177004703) that 
if $\ell(x)$ is the local time at $x$ for Brownian motion run for time 1 and begun at 0,
i.e.,
$$
\begin{align*}
    \ell(x) = \lim_{\epsilon \to 0} \frac{1}{\epsilon}
        \text{measure}\{t \,:\, x \le B_t \le x+\epsilon, \, 0 \le t \le 1\}
\end{align*}
$$
then
$$
\begin{align*}
    \P\{ \ell(x) \le u \} = 2 \Phi(|x|+u)-1,
\end{align*}
$$
where $\Phi(x)=\P\{B_1 \le x\}$ is the Gaussian CDF.
Differentiating this, $\P\{\ell \in du\} = \partial_u \P\{\ell \le u\}$, so
$$
\begin{align*}
    \P\{ \ell(x) = 0 \} &= 2 \Phi(|x|)-1 \\
    \P\{ \ell(x) \in du \} &= 2 \phi(|x|+u) ,
\end{align*}
$$
where $\phi$ is the Gaussian density.

@@

Now suppose that something happens at rate $r$ relative to the local time at $x$
(e.g., recombination to the other background).
Then the chance that this has not happened by time 1 is
$$
\begin{align*}
\E\left[ e^{-r\ell(x)} \right]
&=
2\Phi(|x|)-1 + 
\int_0^\infty e^{-rt} 2 \phi(|x|+t) dt \\
&=
2\Phi(|x|)-1 + 
e^{r|x|} \int_{|x|}^\infty e^{-rt} 2 \phi(t) dt \\
&=
2\Phi(|x|)-1 + 
2 e^{r|x|} \int_{|x|}^\infty \frac{1}{\sqrt{2\pi}} \exp\left( - rt - t^2/2 \right) dt \\
&=
2\Phi(|x|)-1 + 
2 e^{r|x|+r^2/2} \int_{|x|}^\infty \frac{1}{\sqrt{2\pi}} \exp\left( - (t+r)^2/2 \right) dt \\
&=
2\Phi(|x|)-1 + 
2 e^{r|x|+r^2/2} \left( 1-\Phi(|x|+r) \right) .
\end{align*}
$$
At $r=0$ this equals 1, as it should, since it's the chance that nothing happens.
For large $x$, since $1-\Phi(x) \sim e^{-x^2/2}/\sqrt{2\pi}x$,
this is approximately 
$1 + 2 \exp( r|x| + r^2/2 - (|x|+r)^2/2 )$,
which goes to 1, as it should.

@@

This is very steep at $x=0$.
Notice that as $x$ goes to 0,
$$
\begin{align*}
    \lim_{x \to 0} \zeta(x,r) 
    &= 2 e^{r^2/2} (1-\Phi(r)) \\
    &= \frac{2}{r\sqrt{2\pi}} + O(r^{-3}) .
\end{align*}
$$
so that the difference in the left and right limits is
$$
\begin{align*}
    \lim_{x \to 0-} p(x,r) 
    -\lim_{x \to 0+} p(x,r) 
    &= 
    \frac{1}{2}+\frac{2}{r\sqrt{2\pi}} 
    - \frac{1}{2}-\frac{2}{r\sqrt{2\pi}} 
      +O(r^{-3}) \\
    &=
    \frac{4}{r\sqrt{2\pi}}
      +O(r^{-3}) 
\end{align*}
$$

Let's also look at the slope of $p$ at $x=0$.
Since $\partial_x |x| = \sgn(x)$,
$$
\begin{align*}
    \partial_x \zeta(x,r)
    &=
    2 \sgn(x) \left(
    \phi(|x|)
    + r e^{r|x|+r^2/2} (1-\Phi(|x|+r))
    - e^{r|x|+r^2/2} \phi(|x|+r)
      \right)
\end{align*}
$$
and so
$$
\begin{align*}
    -\partial_x p(x)
    &=
    \phi(|x|)
    + r e^{r|x|+r^2/2} (1-\Phi(|x|+r))
    - e^{r|x|+r^2/2} \phi(|x|+r) \\
    &\to
    \frac{1}{\sqrt{2\pi}}(1-e^{r^2/2}e^{-r^2/2})
    + r e^{r^2/2} (1-\Phi(r))  \qquad \text{as } x \to 0 \\
    &=
    r e^{r^2/2} (1-\Phi(r)) \\
    &=
    \frac{1}{\sqrt{2\pi}} + O(\frac{1}{r}) .
\end{align*}
$$

Also note that for large $|x|$, by the asymptotics for $1-\Phi(x)$,
$$
\begin{align*}
    \zeta(x,r) = 1 - \sqrt{\frac{2}{\pi}} e^{-x^2/2} \left( \frac{r}{x+r} \right) .
\end{align*}
$$

@@

The key quantity we want is the probability that a Brownian motion with variance $\sigma^2$ killed at rate $r$
with respect to its local time at $x$ was still alive at time $\tau$.
Denote this by
$$
\begin{align*}
    \zeta(x,r,\tau,\sigma).
\end{align*}
$$
By Brownian scaling,
since $B_{ct}$ has the same distribution as $\sqrt{c}B_t$,
$$
\begin{align*}
    \zeta(x,r,\tau,\sigma) 
    &=
    \zeta(x,r\tau,1,\sqrt{\tau}\sigma) \\
    &=
    \zeta(x/\sqrt{\tau\sigma^2},r\tau) ,
\end{align*}
$$
where
$$
\begin{align*}
    \zeta(x,r)
    &=
    1 - 2(1-\Phi(|x|)) + 
    2 e^{r|x|+r^2/2} \left( 1-\Phi(|x|+r) \right) .
\end{align*}
$$

@@

Now, let $p(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
Let $p_H$ denote the density of heterozygotes in the cline
-- this should be $p_H = (1-s)/2$).
For an individual to the left of the barrier ($x<0$),
the probability of inheriting from the left of the barrier at $\tau$
is the probabiltiy that no recombination at the barrier has occurred,
plus one-half the probability that recombination in a heterozygote has occurred,
i.e., $\zeta p_H +(1-\zeta p_H)/2=(1+\zeta p_H)/2$.
If $x>0$ this is the one-half times probability a recombination in a heterozygote has occurred,
i.e., $(1-\zeta p_H)/2$.
Combining these,
the probability of ending up on the left-hand side is
$$
\begin{align*}
    p(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,r,\tau) p_H \right\} .
\end{align*}
$$

>>>>

It states in [Takacs](https://projecteuclid.org/euclid.aoap/1177004703) that 
if $\ell(x)$ is the local time at $x$ for Brownian motion run for time 1 and begun at 0,
i.e.,
$$
\begin{aligned}
    \ell(x) = \lim_{\epsilon \to 0} \frac{1}{\epsilon}
        \text{measure}\{t \,:\, x \le B_t \le x+\epsilon, \, 0 \le t \le 1\}
\end{aligned}
$$
then
$$
\begin{aligned}
    \P\{ \ell(x) \le u \} = 2 \Phi(|x|+u)-1,
\end{aligned}
$$
where $\Phi(x)=\P\{B_1 \le x\}$ is the Gaussian CDF.
Differentiating this, $\P\{\ell \in du\} = \partial_u \P\{\ell \le u\}$, so
$$
\begin{aligned}
    \P\{ \ell(x) = 0 \} &= 2 \Phi(|x|)-1 \\
    \P\{ \ell(x) \in du \} &= 2 \phi(|x|+u) ,
\end{aligned}
$$
where $\phi$ is the Gaussian density.

@@

Now suppose that something happens at rate $r$ relative to the local time at $x$
(e.g., recombination to the other background).
Then the chance that this has not happened by time 1 is
$$
\begin{aligned}
\E\left[ e^{-r\ell(x)} \right]
&=
2\Phi(|x|)-1 + 
\int_0^\infty e^{-rt} 2 \phi(|x|+t) dt \\
&=
2\Phi(|x|)-1 + 
e^{r|x|} \int_{|x|}^\infty e^{-rt} 2 \phi(t) dt \\
&=
2\Phi(|x|)-1 + 
2 e^{r|x|} \int_{|x|}^\infty \frac{1}{\sqrt{2\pi}} \exp\left( - rt - t^2/2 \right) dt \\
&=
2\Phi(|x|)-1 + 
2 e^{r|x|+r^2/2} \int_{|x|}^\infty \frac{1}{\sqrt{2\pi}} \exp\left( - (t+r)^2/2 \right) dt \\
&=
2\Phi(|x|)-1 + 
2 e^{r|x|+r^2/2} \left( 1-\Phi(|x|+r) \right) .
\end{aligned}
$$
At $r=0$ this equals 1, as it should, since it's the chance that nothing happens.
For large $x$, since $1-\Phi(x) \sim e^{-x^2/2}/\sqrt{2\pi}x$,
this is approximately 
$1 + 2 \exp( r|x| + r^2/2 - (|x|+r)^2/2 )$,
which goes to 1, as it should.

@@

This is very steep at $x=0$.
Notice that as $x$ goes to 0,
$$
\begin{aligned}
    \lim_{x \to 0} \zeta(x,r) 
    &= 2 e^{r^2/2} (1-\Phi(r)) \\
    &= \frac{2}{r\sqrt{2\pi}} + O(r^{-3}) .
\end{aligned}
$$
so that the difference in the left and right limits is
$$
\begin{aligned}
    \lim_{x \to 0-} p(x,r) 
    -\lim_{x \to 0+} p(x,r) 
    &= 
    \frac{1}{2}+\frac{2}{r\sqrt{2\pi}} 
    - \frac{1}{2}-\frac{2}{r\sqrt{2\pi}} 
      +O(r^{-3}) \\
    &=
    \frac{4}{r\sqrt{2\pi}}
      +O(r^{-3}) 
\end{aligned}
$$

Let's also look at the slope of $p$ at $x=0$.
Since $\partial_x |x| = \sgn(x)$,
$$
\begin{aligned}
    \partial_x \zeta(x,r)
    &=
    2 \sgn(x) \left(
    \phi(|x|)
    + r e^{r|x|+r^2/2} (1-\Phi(|x|+r))
    - e^{r|x|+r^2/2} \phi(|x|+r)
      \right)
\end{aligned}
$$
and so
$$
\begin{aligned}
    -\partial_x p(x)
    &=
    \phi(|x|)
    + r e^{r|x|+r^2/2} (1-\Phi(|x|+r))
    - e^{r|x|+r^2/2} \phi(|x|+r) \\
    &\to
    \frac{1}{\sqrt{2\pi}}(1-e^{r^2/2}e^{-r^2/2})
    + r e^{r^2/2} (1-\Phi(r))  \qquad \text{as } x \to 0 \\
    &=
    r e^{r^2/2} (1-\Phi(r)) \\
    &=
    \frac{1}{\sqrt{2\pi}} + O(\frac{1}{r}) .
\end{aligned}
$$

Also note that for large $|x|$, by the asymptotics for $1-\Phi(x)$,
$$
\begin{aligned}
    \zeta(x,r) = 1 - \sqrt{\frac{2}{\pi}} e^{-x^2/2} \left( \frac{r}{x+r} \right) .
\end{aligned}
$$

@@

The key quantity we want is the probability that a Brownian motion with variance $\sigma^2$ killed at rate $r$
with respect to its local time at $x$ was still alive at time $\tau$.
Denote this by
$$
\begin{aligned}
    \zeta(x,r,\tau,\sigma).
\end{aligned}
$$
By Brownian scaling,
since $B_{ct}$ has the same distribution as $\sqrt{c}B_t$,
$$
\begin{aligned}
    \zeta(x,r,\tau,\sigma) 
    &=
    \zeta(x,r\tau,1,\sqrt{\tau}\sigma) \\
    &=
    \zeta(x/\sqrt{\tau\sigma^2},r\tau) ,
\end{aligned}
$$
where
$$
\begin{aligned}
    \zeta(x,r)
    &=
    1 - 2(1-\Phi(|x|)) + 
    2 e^{r|x|+r^2/2} \left( 1-\Phi(|x|+r) \right) .
\end{aligned}
$$

@@

Now, let $p(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
Let $p_H$ denote the density of heterozygotes in the cline
-- this should be $p_H = (1-s)/2$).
For an individual to the left of the barrier ($x<0$),
the probability of inheriting from the left of the barrier at $\tau$
is the probabiltiy that no recombination at the barrier has occurred,
plus one-half the probability that recombination in a heterozygote has occurred,
i.e., $\zeta p_H +(1-\zeta p_H)/2=(1+\zeta p_H)/2$.
If $x>0$ this is the one-half times probability a recombination in a heterozygote has occurred,
i.e., $(1-\zeta p_H)/2$.
Combining these,
the probability of ending up on the left-hand side is
$$
\begin{aligned}
    p(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,r,\tau) p_H \right\} .
\end{aligned}
$$
